Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-01-18 LaTeX/S3_Appendix.tex
moved to plos template, appendices have references that are still messed up

\title{Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems}
\author{Michael F Scott$*^1$, Matthew M Osmond$*^2$, and Sarah P Otto$^2$}
\date{}
\maketitle
\noindent
$*$ These authors contributed equally to this work

\noindent
$^1$ Department of Botany, University of British Columbia, \#3529 - 6270 University Boulevard, Vancouver, BC, Canada V6T 1Z4

\noindent
$^2$ Department of Zoology, University of British Columbia, \#4200 - 6270 University Boulevard, Vancouver, BC, Canada V6T 1Z4

\noindent
email: mfscott@biodiversity.ubc.ca, mmosmond@zoology.ubc.ca

\noindent
Contributions:

\newpage
\linenumbers
\modulolinenumbers[2]

\begin{abstract}
Sex-determination systems are remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions %in which new sex-determining systems spread by selection 
%involve sex ratio selection or sex-specific selection (e.g., sexually antagonistic selection).
%These studies 
generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a sexually-antagonistic locus. 
We use population genetic models to extend these theories in two ways: 
(1) We explicitly consider selection on loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
%(1) We explicitly consider how selection on very tightly sex-linked loci influences the spread of novel sex-determiners. 
Variation at such loci can favour the spread of new sex-determination systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or unlinked) to the locus under selection, which is not predicted by previous theory. 
%With tight linkage, selection can maintain equilibria at which the allele fixed on the sex-limited chromosome (Y or W) is simultaneously favoured in the homogametic sex (XX females or ZZ males) such that heterogametic transitions (XY to ZW or ZW to XY) are favoured by selection. 
%Alternatively, a neo-W (neo-Y) locus can invade when it allows greater specialism on alleles beneficial in females (male) compared to an ancestral XX female (ZZ male), 
(2) We also consider selection upon haploid genotypes either during gametic competition (e.g., pollen/sperm competition) or meiosis (i.e., non-Mendelian segregation); selective processes that typically occur in one sex or the other. 
%As well as having sex-specific fitness consequences, haploid selection can cause the zygotic sex ratio to become biased because sex ratios are determined by the production and fertilization success of X- versus Y-bearing pollen/sperm (or Z- versus W-bearing ovules/eggs). 
%Consequently, selection for XY to ZW transitions and ZW to XY transitions can be assymetrical when linkage between the ancestral sex-determining locus and a locus under haploid selection is tight, in which case ancestral sex ratio biases can be strong.
%In addition, when linkage between the ancestral sex-determining locus and a locus under haploid selection is tight, sex ratio biases can be strong such that there is an asymmetry between invasion of ancestrally XY and ancestrally ZW systems because, e.g., haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system. 
We find that associations with haploid selected loci can drive transitions between sex determination systems, without requiring sexually-antagonistic selection in diploids. 
Unexpectedly, with haploid selection, transitions between male and female heterogamety can also evolve where linkage with the sex-determining locus is weakened. 
Furthermore, haploid selection in the heterogametic sex can cause sex ratio biases, which may increase or decrease with the spread of new sex chromosomes. 
Thus, we find that transitions between sex-determination systems cannot be simply predicted by selection to equalise the sex ratio. 
%These results indicate that loci experiencing haploid selection can favour the evolution of new sex-determination systems. 
%These results indicate that favourable associations that develop between the ancestral sex-determining locus and selected loci can be broken during the spread of a new sex-determining region. 
%Such transitions are not possible with diploid selection alone, in which case tighter linkage increases the fitness of both males and females. 
%Notably, we find that the spread of new genetic sex-determination systems is not affected by sex ratio biases that are caused by haploid selection. 
%A surprising result given that other determinants of sex allocation typically experience strong Fisherian sex ratio selection to equalize sex ratios. 
%Overall, we demonstrate when and how the evolution of sex determining systems is impacted by sex ratio biases resulting from haploid selection and we show that new sex-determining regions can be favoured even if they are not closely linked with a locus under selection. 
Overall, our models reveal that transitions between sex-determination systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
These results predict conditions under which sex-determination systems are likely to be labile and draw novel connections with sex ratio evolution
%These results might be reflected in the lability with which sex-determination systems evolve. %and for the factors determining the phylogenetic pattern of sex-determination systems. 
\end{abstract}

%abstract word count: $\approx$ 350

\newpage

\section*{Introduction}

%CHECK THROUGHOUT: use of sex-determination SYSTEMS versus MECHANISMS. 
%CHECK THROUGHOUT: use of haploid selection, gametic selection, and meiotic drive.

Animals and angiosperms exhibit extremely diverse sex-determination systems \citep[reviewed in][]{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}. 
Among species with genetic sex determination of diploid sexes (GSD), some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? I think monotremes have XY but tonnes of translocations, giving: X1,X2...X6 Y1
mammals and most dioecious plants \citep{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining region changes. 
For example, transitions of the master sex-determining gene between chromosomes or the evolution of new master sex-determining genes have occurred in Salmonids \citep{Li:2011fm,Yano:2012di}, Diptera \citep{Vicoso:2015hf}, and \textit{Oryzias} \citep{Myosho:2012fv}.
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many clades exhibit transitions between male (XY) and female (ZW) heterogamety, including snakes \citep{Gamble2017}, lizards \citep{Ezaz:2009tk}, eight of 26 teleost fish families \citep{Mank:2006bt}, true fruit flies \citep[Tephritids,][]{Vicoso:2015hf}, amphibians \citep{Hillis:1990gu}, the angiosperm genus \textit{Silene} \citep{Slancarova:2013dq}, the angiosperm family \textit{Salicaceae} \citep{Pucholt2015,Pucholt2017} and Coleoptera and Hemiptera \citep[][plate 2]{Beukeboom:2014vb}.
Indeed, in some cases, both male and female heterogametic sex-determination systems can be found in the same species, as reported in houseflies \citep{Macdonald1978}, midges \citep{Thompson1971}, frogs \citep{Ogata:2007jm}, cichlid fish \citep{Ser:2010iq}, tilapia \citep{Lee2004}, sea bass \citep{Vandeputte2007}, and lab-strains of Zebrafish \citep{Liew2012,Wilson2014}.
In addition, multiple transitions have occurred between genetic (GSD) and environmental sex-determination (ESD) systems, e.g., in reptiles and fishes \citep{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

%NOTE: ``Bull & Charnov (1977) hypothesized that a new sex-determining gene can rapidly increase and become fixedin a population if it is linked to a gene with high adaptivevalue, and finally cause a change of the heterogametic sex.'' The Bull and Charnov paper is also the one in which the set of neutral equilibria between XY and ZW are identified (where sex ratios are equal)
%NOTE: Vuilleumier et al. 2007 also find polymorphic sex determination can be stable.

%For example, closely related species in the Medaka lineage (cite Takenhana et al. 2008) and Tilapiinae (cite Lee et al 2004, Cnaani et al 2008), citations in Beukeboom.

%\textcolor{red}{Brief description of sex ratio adjustment and sexual antagonism theories:}

Predominant theories accounting for the spread of new sex-determination systems by selection involve fitness differences between sexes (e.g., sexually antagonistic selection) or sex ratio selection \citep{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
\citet{vanDoorn:2007eu,vanDoorn:2010hu} and \citet{Muralidhar2018} have shown that new sex-determining loci can be favoured if they arise in close linkage with a locus that experiences sexual antagonism. 
Tighter linkage allows a stronger favourable association to build up between a male-beneficial allele, and a neo-Y chromosome, for example. 
Such associations can favour a new partially-masculinizing or partially-feminizing allele \citep{Muralidhar2018}, a new master sex-determining gene \citep{vanDoorn:2007eu}, and transitions between male and female heterogamety \citep[trans-GSD transitions, ZW to XY or XY to ZW,][]{vanDoorn:2010hu}. 
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determination locus will develop similar, favourable associations and are expected to hinder the spread of a new sex-determination system. 
%Here we extend these studies by explicitly calculating the the equilibrium allele frequencies of loci that are very tightly linked to the ancestral sex-determining region.

The sex ratio is directly determined by the sex-determination system, and it has therefore been suggested that sex ratio selection is a dominant force in the evolution of sex determination (e.g., \citealt{Bull:1983vi}, p 66-67; \citealt{Beukeboom:2014vb}, Chapter 7). 
`Fisherian' sex ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce \citep{Fisher:1930wy,Charnov:1982wg}.
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female \citep{West:2009we}. 
Thus, if the population sex ratio is biased towards one sex, the average per-individual contribution of genetic material to the next generation from the opposite sex is greater. 
Therefore, a mutant that increases investment in the rarer sex will spread via the higher per-individual contributions made by that sex. 
%The default mode of sex ratio selection is `Fisherian' sex ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
%Given that the sex-determination system can directly affect the sex ratio, we might expect Fisherian sex ratio selection to influence the spread of new sex-determination systems. 
In the case of sex-chromosome evolution, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased \citep{Hamilton:1967ts}. 
They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations, i.e., neo-Y or neo-W loci) can then spread, which restore an even sex ratio.

%\textcolor{red}{We add haploid selection:}
Here we use mathematical models to find the conditions under which new sex-determination systems spread when individuals experience selection at both diploid and haploid stages. 
Even in animal and plant species that have much larger and more conspicuous diploid phases than haploid phases, many loci experience significant haploid selection through gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-Mendelian) segregation of genotypes during gamete production (from one parent) and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool (potentially from multiple parents); the term `haploid selection' encompasses both processes.

% from Bai et al. (2016) ``[Segregation distortion] phenomenon is commonly detected via genetic mapping and has been documented in various species, including mouse (Eversley et al., 2010; Casellas et al., 2012), Drosophila (Phadnis and Orr, 2009; Larracuente and Presgraves, 2012; McDermott and Noor, 2012), Tigriopus (Pritchard et al., 2011), rice (Koide et al., 2012; Reflinur et al., 2014; Xu et al., 2014), maize (Tang et al., 2013), and cotton (Yu et al., 2011; Hulse-Kemp et al., 2015).'' Also more in that paper that might be useful. 
%Drosophila http://www.biorxiv.org/content/early/2017/02/06/106336
Segregation distortion provides putative evidence of haploid selection and can sometimes be attributed to meiotic drive and/or gametic competition \citep{Lalanne2004,Fishman2005,Leppala2008,Leppala2013,Didion2015,Didion2016}.
Where it has been characterized, meiotic drive generally occurs either during the production of male or female gametes only \citep{Ubeda:2005gw,Lindholm:2016cw}.
Gametic competition is also typically sex specific, occurring primarily among male gametes, because there are typically many more pollen/sperm than required for fertilization.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte, and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Nevertheless, recent studies have demonstrated that sperm competition in animals can alter haploid allele frequencies and increase offspring fitness \citep{Immler:2014im,Alavioon2017}.

There are various ways by which genes experiencing haploid selection could influence transitions between sex-determination systems. 
If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection would affect transitions between sex-determination systems in a similar manner to sex-specific diploid selection \citep[as explored by][]{vanDoorn:2007eu,vanDoorn:2010hu}. 
That is, new masculinizing mutations (neo-Y chromosomes) could be favoured via associations with alleles that are beneficial in the male haploid stage. 
On the other hand, sex ratios can also become biased by linkage between the sex-determining region and a locus that harbours genetic variation in haploid fitness. 
For example, there are several known cases of sex ratio bias caused by sex-linked meiotic drive alleles \citep[][Chapter 3]{Burt:2006} or selection among X- and Y-bearing pollen \citep{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
It is not immediately clear how the spread of new sex-determination systems would be influenced by the combination of sex ratio biases and associations between haploid selected loci and sex-determining regions.

%Perhaps longer discussion of tight sex-linkage here? e.g., mention that r=0 is a particularly significant case for sex chromosomes.

We find that the spread of novel sex-determiners is influenced by both Fisherian sex ratio selection and by selection on genetically-associated alleles. 
%We find that sex ratio biases caused by haploid selection can exert Fisherian sex ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
Surprisingly, Fisherian sex ratio selection does not dominate; it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing increasingly biased sex ratios. 
In addition to considering haploid selection, another novel development in our model is that we consider loci that are in very tight linkage with the ancestral sex-determining region. 
Because sex-determining loci are often found within a region of suppressed recombination, there can be a significant number of tightly linked loci. 
We find that loci linked with the ancestral sex-determining region can drive transitions in which the heterogametic sex changes, even when the neo-sex-determining locus is less closely linked to loci under selection (either including haploid selection or not). 
%Such transitions are not favoured in models lacking tight linkage and/or haploid selection. 
%Secondly, we allow sex-specific haploid selection to occur on a locus in tight or loose linkage with the ancestral sex-determining region. 
%In addition, when linkage between the ancestral sex-determining locus and a locus under haploid selection is tight, sex ratio biases can be strong such that there is an asymmetry between invasion of ancestrally XY and ancestrally ZW systems because, e.g., haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system.

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Model}
%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Model Outline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=\linewidth]{Sex_determination_outline.pdf}}
\caption{
Outline of model features. 
Panel A: Recombination rate parameters between the ancestral-sex-determining locus (here, assumed to have X or Y alleles), a locus under selection (\textbf{A}, with alleles $A$ and $a$), and a neo-sex-determining locus (\textbf{M}, with alleles $M$ and $m$). If $r<1/2$, then associations between ancestral-sex-determining alleles (X and Y) and \textbf{A} locus alleles can be maintained past recombination in males. 
Panel B: Haploid selection is often sex-specific, occurring during haploid production or competition in either males or females. 
For example, haploid selection in males only is represented by the dashed circle.  
If X or Y alleles remain associated with alleles that experience haploid selection in males ($r<1/2$), then zygotic sex ratios can become biased because either X or Y male gametes/gametophytes will be abundant after haploid selection. 
However, the zygotic sex ratio is not biased by male haploid selection in ZW sex-determination systems. 
Similarly, zygotic sex ratio biases can occur if haploid selected alleles are associated with neo-sex-determining alleles ($M$ and $m$, i.e., if $R<1/2$). 
Panel C: During cis-GSD transitions (XY to XY or ZW to ZW, without loss of generality we assume ancestral XY sex determination here), a neo-Y allele spreads to pseudo-fixation (its maximum frequency among male gametes) and the the ancestral-Y allele is lost. 
Panel D: During trans-GSD transitions (XY to ZW or ZW to XY), a neo-W allele spreads to pseudo-fixation (its maximum frequency among female gametes) and the ancestral-X allele is lost. Neo-W mutations allow Y-associated alleles into females, which may impede or aid their spread. 
}
\label{fig:model_outline}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\textcolor{red}{Change all  $\alpha^\Hermaphrodite$ to $(1+\alpha_{\Delta}^\Hermaphrodite)$.}
%\textcolor{blue}{I've attempted this everywhere except in the recursions (S.1), which seem more natural with $\alpha$'s. I've run into trouble in equation S.8c,d -- I think we should check the Mathematica results to be sure we haven't made a typo. This also introduced an extra 1/2 in S.6c,d that might need to be explained.}
%\textcolor{red}{hmm, not sure. This was an idea from Sally, I think in response to terms like $2 \alpha^\Hermaphrodite$ and $2(1-\alpha^\Hermaphrodite)$. It's possible that it makes other equations less easy to understand. My previous (not explained) logic was to use $w$ and $\alpha$ for strong selection and $s$, $t$, and $\alpha_\Delta$ for weak selection. Just to check, I should have written $\alpha^\Hermaphrodite=(1+\alpha_{\Delta}^\Hermaphrodite)/2$, maybe that's where the factors of $1/2$ come from?}
%\textcolor{blue}{shoot, i may have changed too much (i.e. strong selection too), but i did use the correct transformation. still, the form of S.8c,d seems wrong. I'll double check it. CANT FIND ANY ERRORS YET STILL HAVE A "2" IN S.8c. These 2s reappear in S.9 and S.10, always with $W_{AA}$.}

%\textcolor{red}{Switch between $\rho$ and $\rho$ in all places because $\rho$ is used for double recombination events.}
%\textcolor{blue}{I think we need a different variable given the $\rho$ terms in the characteristic polynomial. Also, where is $\rho$ used for double recombination? In other papers? Here $\rho$ is the probability of an odd number of cross-overs between the SDR and M loci.}
%\textcolor{red}{Good point, not sure what's best for \textbf{the-term-formerly-known-as-$\rho$}. Yeah, Sally's written comment was something like ``$\rho$ is semi-standard notation for double recombination so it's use as a recombination rate might be confusing''.}

%\textcolor{red}{Change $\zeta$ to represent zygotic sex ratio of males, consistent with $q$ and figures.}

%genetic system
We consider transitions between ancestral and novel sex-determining systems using a three-locus model, each locus having two alleles. 
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X}; $XX$ genotypes become females and $XY$ become males (or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is epistatically dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, we can interpret $m$ as an environmental sex determination (ESD) allele, such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
%We also analyze a model of maternally-controlled environmental sex-determination, where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in the Appendix. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which assumes a polygamous mating system. 
Gametic competition in monogamous mating systems is, however, equivalent to meiotic drive in our model (described below), as either only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci. %(and the \textbf{M} genotype of their mother in the case of maternal control).
Diploid males and females then experience selection, with relative fitnesses $w_{AA}^{\Hermaphrodite}$, $w_{Aa}^{\Hermaphrodite}$, and $w_{aa}^{\Hermaphrodite}$. %where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Figure \ref{fig:model_outline}A and Table \ref{tab:chisubstitutions}). 
Individuals that are heterozygous at the \textbf{A} locus may experience meiotic drive; a gamete produced by $Aa$ heterozgotes of sex $\Hermaphrodite$ bears allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

\begin{table}[ht]
\centering
\smallskip
\caption{Relative fitness of different genotypes in sex $\Hermaphrodite \in \{\female,\male\}$}
\begin{tabular}{l c }
\hline\hline
  Genotype & Relative fitness during gametic competition \\ [0.5ex] \hline
  A & $w_{A}^\Hermaphrodite = 1+t^\Hermaphrodite$ \\
  a & $w_{a}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Relative fitness during diploid selection \\ [0.5ex] \hline
  AA & $w_{AA}^\Hermaphrodite = 1+ s^\Hermaphrodite$ \\
  Aa & $w_{Aa}^\Hermaphrodite = 1+h^\Hermaphrodite s^\Hermaphrodite$ \\
  aa & $w_{aa}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Transmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  A & $\alpha^\Hermaphrodite=1/2+\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  a & $1-\alpha^\Hermaphrodite=1/2-\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

%Info moved above
%what were going to do, briefly
%With allele $M$ fixed at the \textbf{M} locus, sex is determined by locus \textbf{X} and an equilibrium is reached at locus \textbf{A}. %where, importantly, locus \textbf{A} can be stably polymorphic. %the dynamics can be described by the frequency of $A$ in gametes/gametophytes from the homogametic sex and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex.
%We then examine the ability of a rare, novel sex-determiner, $m$, to increase in frequency from this equilibrium.
%Numerical simulations are used to examine when an invading mutation goes to fixation.

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%

The model outlined above describes both ancestrally-XY and ancestrally-ZW sex-determination systems if we relabel the two sexes as being ancestrally `heterogametic' or ancestrally `homogametic'. 
Without loss of generality, we primarily refer to the ancestrally heterogametic sex as male and the ancestrally homogametic sex as female.
That is, we describe an ancestral XY sex-determination system but our model is equally applicable to an ancestral ZW sex-determination system (relabelling the ancestrally-heterogametic sex as female and the ancestrally-homogametic sex as male and switching the labels of males and females throughout).

%\subsection*{Turnover between sex-determination systems}

\subsubsection*{Generic invasion by a neo-Y or neo-W}

The evolution of a new sex-determination system requires that a rare mutant allele at the novel sex-determining locus, $m$, increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially in a number of cases of interest. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males or by females, respectively. 
Furthermore, if the $m$ allele is fully epistatically dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves the quadratic, $\lambda^2+ b \lambda + c =0$ \textcolor{blue}{(see Appendix for a discussion of other roots - or Sally's proof!)}.
Here, $b= - (\lambda_{mA} + \lambda_{ma})+(\chi_{mA} + \chi_{ma})$ and $c = (\lambda_{mA}-\chi_{mA}) (\lambda_{ma}-\chi_{ma}) -\chi_{mA} \chi_{ma}$, where $\lambda_{mi}$ is the multiplicative growth rate (which we will call the ``haplotypic growth rate'') of the neo-sex determination allele $m$ on background $i$ without accounting for loss due to recombination, and $\chi_{mi}$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other \textbf{A} locus background in heterozygotes (see Table \ref{tab:haplotype_growth}).
The $\lambda_{mi}$ and $\chi_{mi}$, and thus the spread of the mutant $m$ allele, depend on the frequency of alleles at the \textbf{A} and \textbf{X} loci in the ancestral population. 
In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We will consider only equilibrium frequencies of alleles, $\hat{p}^\Hermaphrodite_i$, and Y-bearing male gametes, $\hat{q}$, when calculating the eigenvalues.

%\textcolor{red}{I now remove the zygotic sex ratio $\zeta$ from the mean fitnesses. The mean fitnesses will have to be adjusted in a corresponding way.}
%
%\textcolor{red}{CHECK: DO THE HAPLOID MEAN FITNESSES HAVE TO BE IN THE DENOMINATOR? i think so, based on definitions in table S.2 (mmo)}

% as there are now two 2s in the lambdas of table 2. perhaps the 2 in the denominator is needed because the frequency of males sums to one as does the frequency of females?}

\begin{threeparttable}[ht]
\centering
\smallskip
%\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\caption{Parameters determining invasion of mutant neo-Y and neo-W alleles into an ancestrally XY system}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   $m$ is a neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{YA} = \left( 2 \zeta \right)^{-1} \left[\hat{p}_X^\female w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male} + (1-\hat{p}_X^\female) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1+\alpha_{\Delta}^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^\male \right) $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{Ya} = \left( 2 \zeta \right)^{-1} \left[(1-\hat{p}_X^\female) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male} + \hat{p}_X^\female w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1 - \alpha_{\Delta}^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^\male \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{YA} = R \left( 2 \zeta \right)^{-1} \left[ (1-\hat{p}_X^\female) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1+\alpha_{\Delta}^{\male}) \right]/  \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^\male \right)   $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{Ya} = R \left( 2 \zeta \right)^{-1} \left[   \hat{p}_X^\female w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} (1 - \alpha_\Delta^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^\male \right)  $\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  $m$ is a neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{WA} = \left[ 2 (1 - \zeta) \right]^{-1} \left[ \bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}(1+\alpha_{\Delta}^{\female})\right]/ \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^\female \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{Wa} = \left[ 2 (1 - \zeta) \right]^{-1} \left[ (1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha_{\Delta}^{\female})\right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^\female \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{WA} = R \left[ 2 (1 - \zeta) \right]^{-1} \left[ (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} (1+\alpha_{\Delta}^{\female}) \right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^\female \right)  $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{Wa} = R \left[ 2 (1 - \zeta) \right]^{-1} \left[ \bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} (1-\alpha_{\Delta}^{\female}) \right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^\female \right) $\\ [1ex] 
  \hline \hline 
   \end{tabular}
      \begin{tablenotes}
      \scriptsize
      \item $\bar{p}^{\male}=(1-\hat{q})\hat{p}_X^\male + \hat{q}\hat{p}_Y^\male$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes.
%      \item $R$ is the probability of recombination between loci \textbf{A} and \textbf{M}.
      \item $\zeta$ is the zygotic sex ratio (fraction male)
      \item $\bar{w}^\Hermaphrodite$ is the mean fitness of diploids of sex $\Hermaphrodite$, see Table \ref{tab:meanfitnesses}
      \item $\bar{w}_H^\Hermaphrodite$ is the mean fitness of haploids from sex $\Hermaphrodite$, see Table \ref{tab:meanfitnesses}

>>>>

\setcounter{equation}{0}
\renewcommand{\theequation}{S1.\arabic{equation}}

\section*{S3 Appendix}
